---
title: "Predicting survival rates of passengers in the Titanic"
output: html_notebook
---

```{r Setup}
library(ggplot2)
library(dplyr)
library(readr)
```

```{r Import}
# Let's consolidate the test and train and create one unified dataset
trainDat <- read_csv("train.csv")
testDat <- read_csv("test.csv")
```

```{r}
# Transform several variables into factor data types, used for training and test data
convertToDataType <- function(data){
  data$Sex <- factor(data$Sex)
  data$Pclass <- factor(data$Pclass
      , levels = c(1, 2, 3)
      , labels = c("1st", "2nd", "3rd")
  )
  data$Embarked <- factor(data$Embarked
     , levels = c("C", "Q", "S")
     , labels = c("Cherbourg", "Queenstown", "Southampton")
  )
  data$PClass <- factor(data$Pclass)
  return (data)
}

trainDat <- convertToDataType(trainDat)
testDat <- convertToDataType(testDat)
  
trainDat$Survived <- factor(trainDat$Survived
    , levels = c(0, 1)
    , labels = c("No", "Yes")
)

```


Start with some exploratory data and do some data transformations as needed. 

# Exploratory Data Analysis

```{r}
# Look at age
ggplot(trainDat, aes(Age)) + geom_histogram()
```
```{r}
# Look at fare
ggplot(trainDat, aes(Fare)) + geom_histogram()
ggplot(trainDat, aes(x = 1, y = Fare)) + geom_boxplot()
```

```{r}
# Look at distribution of socio-economic spread
# as expressed with the class of ticket purchased
ggplot(trainDat, aes(x = Pclass)) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous()
```

*55% of tickets bought are for the 3rd class. This is the lowest class on the socio-economic status (SES) indicator.* 

```{r}
# Look at distribution of socio-economic spread
# as expressed with the class of ticket purchased
ggplot(trainDat, aes(x = SibSp)) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous()
```

*More than 65% had no siblings with them on the Titanic.*

```{r}
# Look at distribution of socio-economic spread
# as expressed with the class of ticket purchased
ggplot(trainDat, aes(x = Parch)) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous()
```

*More than 65% traveled with their nannies. Surprising.*

```{r}
# Look at distribution of socio-economic spread
# as expressed with the class of ticket purchased
ggplot(trainDat, aes(x = Embarked)) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous()
```

*65% of the passengers disembarked from Southampton.*

# Binary Classification Analysis

```{r}
  library(caret)
  # Take out name from both train and test
  drops <- c("Name", "Ticket", "Cabin", "PassengerId")
  trainDat <- trainDat[, !(names(trainDat) %in% drops)]
  testDat <-testDat[, !(names(testDat) %in% drops)]
  
  # fit model
  fit <- glm(Survived ~ ., family = binomial, data = trainDat)
  
  # summarize the fit
  summary(fit)
  
  # make predictions
  probabilities <- predict(fit, testDat, type="response")
  
  table(trainDat$Survived, probabilities > 0.5)
```

The most important predictors are PClass (2nd and 3rd class), Sex, Age, and SibSp in this dataset.